---
title: "Data Management and Analysis Workshop: Part 2"
format: pdf
editor: visual
---

# Data analysis

We will walk through some basic data cleaning, analysis, and visualization tasks using per pupil expenditures as a running example. The repository contains all the data and code we need, so you can just follow along! This notebook introduces descriptive statistics and regression.

## Set up

```{r}
# Install required packages.
req <- c("tidyverse", "openxlsx", "broom")
new <- req[!(req %in% installed.packages()[, "Package"])]
if (length(new)) install.packages(new)

# Load required packages.
library(dplyr)
library(tidyr)
library(broom)
library(openxlsx)
```

```{r}
# Identify inputs and outputs.
PWD <- getwd()
DTA <- file.path(PWD, "out", "data.rda")
OUT <- file.path(PWD, "out", "tables.xlsx")
```

```{r}
# Load analysis file.
# This loads the dataset we created into the same namespace.
load(DTA)
head(dta)
```

## Descriptive statistics

```{r}
# Summarize per pupil expenditure and racial composition by majority status.
tab1 <- dta |>
  group_by(maj_group) |>
  summarize(
    n = n(),
    across(ppe, list(mean = mean, min = min, max = max), .names = "{.col}_{.fn}"),
    ppe_p25 = quantile(ppe, probs = 0.25),
    ppe_p50 = quantile(ppe, probs = 0.5),
    ppe_p75 = quantile(ppe, probs = 0.75),
    across(starts_with("pct_"), ~ mean(.x))
  ) |>
  mutate(pct = n / nrow(dta) * 100) |>
  relocate(pct, .after = n) |>
  relocate(c(ppe_p25, ppe_p50, ppe_p75), .after = ppe_min)
```

## RQ1: How does per pupil expenditure vary by school district composition?

```{r}
# Set the reference group among categories.
table(dta$maj_group)
dta$maj_group <- relevel(as.factor(dta$maj_group), ref = "white")
```

```{r}
# Estimate the partial effects of district composition on per pupil expenditure.
f1 <- ppe ~ maj_group
reg1 <- lm(f1, data = dta)
summary(reg1)
```

## RQ2: Do these differences change with the size of the majority group?

```{r}
# Get the size of the majority group, i.e., the largest share.
dta <- dta |>
  rowwise() |>
  mutate(
    pct_max = max(c_across(starts_with("pct_"))),
    pts_maj = pct_max - 50
  ) |>
  ungroup()
```

```{r}
# Estimate the partial effects of group majority size on per pupil expenditure.
f2 <- ppe ~ maj_group * pts_maj
reg2 <- lm(f2, data = dta)
summary(reg2)
```

## Save tables

```{r}
# Write tables to workbook.
# Provide a named list to write dataframes to different worksheets.
ws <- list("Summary" = tab1, "RQ1" = tidy(reg1), "RQ2" = tidy(reg2))
write.xlsx(ws, OUT)
```